The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-07-30 paper/referee/response-revision1.txt

[Authors' response]
The referee questioned the nature of the objects which are marked as "reliable"
in the catalogue, and in particular questioned the nature of the large number of 
sources not flagged as "reliable".  Having considered this comment, we have come 
to conclude that we made a poor choice of terminology by adopting the labels 
"reliable" and "veryReliable" as quality qualifiers.
These classes, which are based on stringent quality criteria set out in 
Tables 6 and 7, aim to help users identify high-quality objects, e.g. by 
demanding a 10-sigma detection in all bands.  It goes without saying that 
there are plenty of objects which do not strictly satisfy these criteria but are 
nevertheless genuine astrophysical objects.  The qualifier "reliable" might 
however confer the incorrect impression that such sources are "unreliable".
As a result, we have decided to change the terminology by renaming the class 
previously known as "reliable" into "a10" (which is short for "all-band 10-sigma"), 
and renaming "veryReliable" into "a10point".  We believe this change in 
terminology will better serve as an operational description of the detection quality.

In addition to this change in terminology, we have adopted the referee's suggestion
to gauge the reliability of our objects against an external catalogue.  That is,
we have introduced a new Section 6.2, titled "Source reliability", in which the 
reliability of IPHAS sources as a function of the quality criteria is gauged 
by comparing out catalogue against SDSS.  Moreover, we note that a separate paper 
is in an advanced state of preparation (Farnhill et al.), which will investigate 
the issues surrounding the completeness and reliability of our catalogue, 
e.g. by injecting artificial sources into the images and measuring their recovery 
rates, whilst also comparing source counts between SDSS and IPHAS in more detail.

[Referee]
> 7. The authors have done a good job describing the photometric measurements and
> calibration accuracy.  However, there is relatively little discussion of the
> astrometric position reconstruction other than saying it is done relative to
> the 2MASS Point Source Catalogue, and there is no validation of the astrometric
> quality of the Catalogue other than to mention matching success.  Some quantification
> of the astrometric performance should be added (e.g. mean position residuals as a function
> of magnitude from comparisons with the astrometric reference stars, or external
> astrometric collections, such as UCAC4).
> Positions in the IPHAS Catalogue have been reconstructed with respect to 2MASS.
> However, the authors do not state if the proper motion of the 2MASS reference stars 
> was taken into account.  Although small, the net motion of the reference stars can
> collectively imprint on the Catalogue because the IPHAS observations were made at
> different epochs than 2MASS, in some cases as much as 12 years later.   
> Failure to do so can imprint the proper motion onto IPHAS, and also because the 
> IPHAS measurements span nearly 10 years (2003-2012), there can be secular variations.

[Authors' response]
We have edited Section 3.1 to offer the reader more details on the astrometric 
calibration.  This section now also includes the results of an external validation of 
our astrometry against UCAC4, which showed a mean residual of 0.09 +/- 0.07 arcsec at the 
bright end of our catalogue.  We have also added a statement to specify that
no proper motion information was used in arriving at our astrometry, which is
of limited effect because hundreds 2MASS positional reference stars were available 
in each CCD to fit the parameters of the WCS solution.

[Referee]
(There does not appear to be an item number 8 in the referee report.)
> 9. The authors point out in the introduction that IPHAS will complement a number of
> new infrared surveys of the Galactic Plane.  The paper would benefit from at least
> one cross-comparison with the catalogue of one of those surveys, such as GLIMPSE
> or MIPSGAL, as an illustration of how the combination of surveys will enable
> new science.  At present, the only tie-in to the infrared surveys is the
> visual comparison between IPHAS and GLIMPSE images of one field.

### 2014-07-30 paper/iphas-dr2-rev1.tex
Improved astrometry section

{ \bf
Object detection and parametrisation is then carried out
using the standard methods in use by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated
by computing the sigma-clipped median intensity
in a grid of 64$\times$64 pixel bins across the image.
Bilinear interpolation is then used to obtain an estimate 
of the background level at each pixel,
which is robust against contaminating sources.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring pixels
in which the intensity exceeds the background level by at least $1.25\sigma$.
\item The object are parametrised:
positions are obtained by computing 
the intensity-weighted isophotal centre-of-gravity
of each object,
and photometry is obtained 
by measuring the intensity
in a series of soft-edged circular apertures of increasing diameter 
(1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6 and 6\farcs 6).
Objects are also classified morphologically
-- stellar, extended or noise --
based on their intensity as a function of aperture size.
\item The parametrisation of overlapping objects
is refined by simultaneous fitting of
soft-edged apertures to each blend,
effectively carrying out profile fitting where necessary.
We note that the parametrisation of these blended objects is naturally
less reliable than that of single, unconfused sources,
but they are clearly flagged in the catalogue
(to be explained in \S\ref{sec:catalogue}).
\end{enumerate}

@@

Having carried out object detection and characterisation,
an astrometric solution is then fitted
based on the Two-Micron All Sky Survey (2MASS)
point source catalogue \citep{Skrutskie2006},
which itself is calibrated 
in the International Celestial Reference System (ICRS).
A provisional photometric calibration is also determined 
based on the average zeropoint
obtained from sets of standard stars observed within the same night.

%%(even though the nominal rms accuracy
%-- estimated against 2MASS across the focal plane --
%tends to be better than 0.1~arcsec).

@@

The pipelined data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a proper source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which distinguish the present release
from the IDR, which enforced less stringent quality limits,
did not offer a global calibration,
and provided a catalogue in which duplicate detections
of unique sources were not flagged.
These new processing steps are explained next.}

@@

(1) \emph{Depth.} 
We discarded any exposures 
for which the $5\sigma$ limiting magnitude\footnote{\bf
We defined the $5\sigma$ limiting magnitude
as the magnitude which a point source would have,
when its flux in a 2\farcs 3-diameter aperture
is five times the level of the noise in the sky background.
The sky noise was estimated using a robust MAD estimator 
for noise scaled to equivalent Gaussian rms value, ie. = MAD x 1.48,
after removing large scale sky background variations.
(MAD = Median of the Absolute Deviations about the median.)}
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha.
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.

>>>>

{ \bf
Object detection and parametrisation is then carried out
using the standard methods in use by CASU,
which can be summarised in four steps
\citep[a discussion on each of these steps
and related points can be found in][]{Irwin1985,Irwin1997}:
\begin{enumerate}
\item The local sky background is estimated
by computing the sigma-clipped median intensity
in a grid of 64$\times$64 pixel bins across the image,
which is robust against contaminating sources.
Bilinear interpolation is then used to obtain an estimate 
of the background level at each pixel.
\item Objects, or blends of objects, are detected
by identifying groups of 4 or more neighbouring pixels
in which the intensity exceeds the background level by at least $1.25\sigma$.
\item The object are parametrised:
positions are obtained by computing 
the intensity-weighted isophotal centre-of-gravity
of each object,
and photometry is obtained 
by measuring the intensity
in a series of soft-edged circular apertures of increasing diameter 
(1\farcs 2, 2\farcs 3, 3\farcs 3, 4\farcs 6 and 6\farcs 6).
Objects are also classified morphologically
-- stellar, extended or noise --
based on their intensity as a function of aperture size.
\item The parametrisation of overlapping objects
is refined by simultaneous fitting of
soft-edged apertures to each blend,
effectively carrying out profile fitting where it is necessary.
We note that the parametrisation of these blended objects is naturally
less reliable than that of single, unconfused sources,
which is why they are flagged in the catalogue
(to be explained in \S\ref{sec:catalogue}).
\end{enumerate}

@@

Having carried out object detection and characterisation,
the astrometric calibration is then determined.
The solution starts with a rough 
World Coordinate System (WCS)
based on the known telescope and camera geometry,
which uses a Zenithal Polynomial projection
\citep[ZPN;][]{Calabretta2002}
to model the field distortion of the WFC.
The parameters of this solution
are then progressively refined using the
Two-Micron All Sky Survey catalogue \citep[2MASS; ][]{Skrutskie2006}.
We did not include proper motion information during this step,
because each CCD typically contains
hundreds of 2MASS reference stars.
The resulting fit has previously been shown
to deliver results which are internally consistent
to better than 0\farcs 1 across the detector
\citep{Gonzalez-Solares2008}.

An external validation has been carried out
by comparing our positions against the
United States Naval Observatory CCD Astrograph Catalog 
\citep[UCAC4;][]{UCAC4}.
Fig.~\ref{fig:astrometry} shows the distribution
of the astrometric residuals
between IPHAS and UCAC4
for N million stars in the magnitude
range $13 < r < 15$,
which is the range where both surveys overlap
and where the formal mean error of UCAC4
is better than 50~mas \citep{UCAC3}.
We find the mean residual to be $0\farcs 09\pm0\farcs 07$,
which is satisfactory for our purposes.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/astrometry/astrometry.pdf} 
    \caption{Distribution of position residuals
            between IPHAS DR2 and UCAC4.
            The residuals were computed for stars
            in the magnitude range $13 < r < 15$ which were
            not blended and not saturated.}
    \label{fig:astrometry}
\end{figure}

%which itself is calibrated
%in the International Celestial Reference System (ICRS).
%(Figure: note that blended stars were excluded).

%A provisional photometric calibration is also determined 
%based on the average zeropoint
%obtained from sets of standard stars observed within the same night.

@@

The entire data set -- comprising 2.5~terabyte of FITS files --
was then transferred to the University of Hertfordshire
for the purpose of transforming the raw
detection tables into a proper source catalogue which
(i) is quality-controlled,
(ii) is homogeneously calibrated, and 
(iii) contains user-friendly columns and warning flags.
It is these post-processing steps which distinguish the present release
from the IDR; which enforced less stringent quality limits,
did not offer a global calibration,
and provided a catalogue in which duplicate detections
of unique sources were not flagged.
These new processing steps are explained next.}

@@

(1) \emph{Depth.} 
We discarded any exposures 
for which the $5\sigma$ limiting magnitude\footnote{\bf
We defined the $5\sigma$ limiting magnitude
as the magnitude which a point source would have,
when its flux in the 2\farcs 3-diameter aperture
equals five times the level of the noise in the sky background.
The sky noise was estimated using a robust MAD estimator 
for noise scaled to equivalent Gaussian standard deviation,
ie. = MAD x 1.48,
after removing large scale sky background variations.
(MAD = Median of the Absolute Deviations about the median.)}
was brighter than 20th magnitude in the $r$-band
or brighter than 19th in $i$ or \ha.
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields;
the mean depths and standard deviations are 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended
most strongly on the presence of the moon,
which was above the horizon during 62 per cent 
of the observations.
The great range in sky brightness this
produced is behind the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).
To a lesser extent, the wide spread in the $r$-band depth
is also explained by the shorter exposure time
that was used for this band during the first months of data-taking.
